---
title: "Introduction to klcluster package"
author: "Natasja van de L'Isle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = '75%',
  out.height = '75%',
  dev = 'svg'
)
```

The klcluster package is specifically developed to cluster trajectories. This 
tutorial will use the example data that is attached to the package to illustrate
how the package works. This data is provided by R. Mann 
[see @mann2010; @mann2014; @mann2016]
In case you would like some more background information on the (k,l)-clsutering
alorithm I would recommend you to read [@kl_unpublished]


## Setup

Before we start you first need to install the package. 

To install the package use `install.packages(klcluster)`.
Do not forget to load it into the environment: `library(klcluster)`
***NOTE: need to change this in order to download from git instead of CRAN***

In case you want to work with data from movebank, also install and load the
"move" package. For a detailed tutorial see [Move package introduction](https://cran.r-project.org/web/packages/move/vignettes/move.pdf)


## Loading data

There are 4 datasets attached to the package from R. Mann. Here we will use the 
dataset "brc", which contains 24 paths. The standard format the package uses is 
the same as for "brc". 

Note that when you are using your own data, ensure that the coordinates are ordered
by timestamp.


```{r, load data}
library(klcluster)

print(getwd())

data(brc)
```

The standard data format consists of a list of data.frames, where each data.frame 
is one path. The data.frames contain two columns, long and lat, and each row 
corresponds to a coordinate of the path.

```{r, show structure data}
str(brc[1:3])

```

## Similarity measures

There are three similarity measures that can be used. Continuous Frechet, discrete
frechet and Dynamic Time Warping(DTW).

```{r, similarity}
# set seed for reproducibility
set.seed(0)

# choose two paths
p1 <- brc[[1]]
p2 <- brc[[2]]

frechet_distance(p1, p2, 0.001) # 0.001 denotes the precision - see documentation
discrete_frechet(p1, p2)
dtw_dist(p1, p2, TRUE, normalize_dist = TRUE)

```

## Simplification

Given that the clustering algorithms will simplify the paths, we will show how this
works. You need to specify the number of data points the simplified path should
have. Furthermore, if you would want to simplify all paths at once use the
function`simplify_all()`. In case you would like to simplify to epsilon, use `simplify_curve()`. The simplification algorithm implemented uses the
continuous frechet distance.

```{r, simplify, fig.show='hold'}

p1_simple <- bin_simplify(path = p1, size = 20, precision = 1)

library(ggplot2)

plot1 <- ggplot() +  geom_path(data = p1, aes(x = lat, y = long))

plot2 <- plot1 +
  geom_path(data = p1_simple, aes(x = lat, y = long), col = "red") +
  geom_point(data = p1_simple, aes(x = lat, y = long), col = "red")

plot2 + theme_classic()

```


## Clustering

The package contains several methods for clustering. The main method we advise you to use is the (k,l)-clustering algorithm as it is the most efficient algorithm.

However, you can also use the single linkage, k-median, k-means or 1-NN algorithm.
Note that k-median and k-means take as input an initial clustering, hence, you will
first need to run either (k,l)-clustering or single linkage first.

Each of the methods returns a set of centers which are simplified to target
value 'l'. Input 'k' denotes the number of clusters we want. It is also possible
to specify the 'radius' in (k,l)-clustering and single linkage. If you omit 'k',
the algorithm will return the number of clusters such that the radius is smaller
than or equal to input 'radius'. However, note that it might be a good idea to
first pick a 'k' to get a sense of the value for 'radius' which equals the cost
that can also be found in the output.

Also note that the 1-NN algorithm only has one solution, given that you always
match a path with its closest neighbor. Moreover, this solution never contains
a cluster containing only one path as each path is mapped to its closest neighbor.

### (k,l)-clustering algorithm

This algorithm is an efficient clustering algorithm. It randomly picks a trajectory
as a first center. Next, it find the trajectory that is furthest away to choose
as the next center. It continues until either 'k' or 'radius' is satisfied.

```{r, kl-cluster}
brc_kl <- k_l_cluster(brc, k = 5, l = 15)

# the output has the following structure:
str(brc_kl)
```

```{r, show access}

# you can access any element as you would in a normal list:
brc_kl$clusters
```

#### Plotting (k,l)-cluster

Plotting the clusters can be done in many ways. The package contains three built
in functions to plot the clusters. Note that these functions require additional
packages (data.table, ggplot2, colorspace, cowplot, ggmap, mapproj, sp).


```{r, plot clustering, fig.show='hide'}
library(data.table)
library(ggplot2)
library(colorspace)
```

```{r, plot kl}
plot_all_clust_grid(brc, brc_kl, "(k,l)-clustering")

```

We can also use the ggmap package [@ggmap_ref] to plot the trajectories
on a map.

```{r, loadmaps-packages, echo = FALSE}
require(ggmap) #these packages are necessary to work with google maps
require(mapproj)
require(sp)
require(raster)
```

```{r, plot_maps}
plot_all_clust_map(brc, brc_kl, "(k,l)-clustering on map", map_type = "terrain")
```


### Single linkage and 1NN

Now we can compare the (k,l)-clustering algorithm to the other methods. First we
use single linkage and 1NN.

The 1NN algorithm can give a rough idea of how the trajectories are clustered.
Each trajectory is mapped to its closest neighbor.

Single linkage builds a hierarchy of clusters. It then picks the clusters,
depending on 'k' or 'radius'. Note that this is not a heuristic algorthm, thus,
it can take some time to run for large datasets.

```{r, compare}

brc_sl <- single_link(brc, k = 5, l = 15)
brc_sl$clusters

brc_nn <- kNN_trajectories(brc, l = 15)
brc_nn$clusters

```

```{r, compare2-plot,  fig.show = 'hold', echo = FALSE}

print(plot_all_clust(brc, brc_sl, "Single linkage clustering"))

print(plot_all_clust(brc, brc_nn, "1NN clustering"))

```


### K-median and K-means

We can also try to improve the centers by using k-median and k-means.We can update specfic clusters or we can update all. The standard number of iterations equals 1, but this can
be changed by setting max_iter. If verbose is set to true, the algorithms will
also output the improvements made on the costs.

#### K-median

```{r, k-median}
brc_kmed <- kl_median(brc, brc_kl, max_iter = 5)

plot_all_clust_grid(brc, brc_kmed, "k-median")

```


#### K-means

```{r, k-means}
brc_kmean <- kl_means(brc, brc_kl, max_iter = 5)

plot_all_clust_grid(brc, brc_kmean, "k-means")

```


### fsp-method

Next to the implemented clustering algorithms, a new method has been developed
to improve the quality of the centers. Where each clustering algorithm uses the
original coordinates from a trajectory as center, the fsp-method can have any
coordinate. Assume we have a clustering. Then for each cluster, the coordinates
are mapped to one coordinate of the corresponding centers. Each center is then
updated to correspond to the mean of all points mapped to this point.

#### fsp - dtw with input original centers from output clustering

```{r, fsp1}

fsp_dtw1 <- fsp_centers(brc, brc_kl, dist_measure = "dtw", dtw_settings = c(1, TRUE, TRUE))

plot_all_clust_grid(brc, fsp_dtw1, "fsp method dtw - use original center")

```

#### fsp - dtw with input original path corresponding to center from output clustering

```{r, fsp2}
fsp_dtw2 <- fsp_centers(brc, brc_kl, dist_measure = "dtw", dtw_settings = c(2, TRUE, TRUE))
plot_all_clust_grid(brc, fsp_dtw2, "fsp method dtw - use original path")

```

#### fsp - dtw with input sampled center

```{r, fsp3}
fsp_dtw3 <- fsp_centers(brc, brc_kl, dist_measure = "dtw", dtw_settings = c(3, TRUE, TRUE))
plot_all_clust_grid(brc, fsp_dtw3, "fsp method dtw - sampled along center to 15% path size")

```

#### fsp - dtw with input center with more points
```{r, fsp4}
fsp_dtw4 <- fsp_centers(brc, brc_kl, dist_measure = "dtw", dtw_settings = c(4, TRUE, TRUE))
plot_all_clust_grid(brc, fsp_dtw4, "fsp method dtw - simplified original path to 10% of size")

```



## References
